---
title: "**Practical Machine Learning - Course Project**"
author: "Zach Zaccaro"
date: "Saturday, August 22, 2015"
output: html_document
---

This project was completed for the Practical Machine Learning class in the Coursera Data Science specialization track.  

The Github repository for this project can be found [here](https://github.com/zzaccaro/MachineLearning_CourseProject).  
  
## Background
(Taken from the project description)

"Using devices such as Jawbone Up, Nike FuelBand, and Fitbit it is now possible to collect a large amount of data about personal activity relatively inexpensively. These type of devices are part of the quantified self movement - a group of enthusiasts who take measurements about themselves regularly to improve their health, to find patterns in their behavior, or because they are tech geeks. One thing that people regularly do is quantify how much of a particular activity they do, but they rarely quantify how well they do it. In this project, your goal will be to use data from accelerometers on the belt, forearm, arm, and dumbell of 6 participants. They were asked to perform barbell lifts correctly and incorrectly in 5 different ways. More information is available from the website here: http://groupware.les.inf.puc-rio.br/har (see the section on the Weight Lifting Exercise Dataset)."  

## The Data

The training data for this project are available here: 

https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv

The test data are available here: 

https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv

The data for this project come from this source: http://groupware.les.inf.puc-rio.br/har  

## Project Goals
(Taken from the project description)

"The goal of your project is to predict the manner in which they did the exercise. This is the "classe" variable in the training set. You may use any of the other variables to predict with. You should create a report describing how you built your model, how you used cross validation, what you think the expected out of sample error is, and why you made the choices you did. You will also use your prediction model to predict 20 different test cases. 

1. Your submission should consist of a link to a Github repo with your R markdown and compiled HTML file describing your analysis. Please constrain the text of the writeup to < 2000 words and the number of figures to be less than 5. It will make it easier for the graders if you submit a repo with a gh-pages branch so the HTML page can be viewed online (and you always want to make it easy on graders :-).
2. You should also apply your machine learning algorithm to the 20 test cases available in the test data above. Please submit your predictions in appropriate format to the programming assignment for automated grading. See the programming assignment for additional details."  

## Reproducibility and Loading Libraries

In order for reproducibility of these results, the following libraries are required: 

```{r, warning=FALSE}
library(caret)
library(randomForest)
```

We also need to set a seed for reproducibility:  

```{r}
set.seed(23)
```

## Partitioning and Preprocessing the Data

First we need to load the training and testing data.

```{r}
trainingUrl <- "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv"
testingUrl <- "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv"

training <- read.csv(url(trainingUrl), na.strings=c("NA","#DIV/0!",""))
testing <- read.csv(url(testingUrl), na.strings=c("NA","#DIV/0!",""))
dim(training)
summary(training$classe)
```

As you can see, there are 19622 records with 160 variables. We will be predicting the **classe** variable.  
  
Now we need to partition the training data set into its own training and testing sets to build, train, and test our model.

```{r}
inTrain <- createDataPartition(training$classe, p=0.7, list = FALSE)
modelTrain <- training[inTrain, ]
modelTest <- training[-inTrain, ]
dim(modelTrain)
```

There are 159 columns that we can use to predict the classe, but we probably don't need that many. We can reduce the amount of columns in the dataset by removing ones with more missing values.

```{r}
train_na_sums = sapply(modelTrain, function(x) {sum(is.na(x))})
table(train_na_sums)
```

As the table shows, there are 60 columns with 0 NA values, and the rest contain almost all NA values. Therefore, we are only going to use the 60 columns with no NA values. We also need to do this with the testing data set.

```{r}
test_na_sums = sapply(modelTest, function(x) {sum(is.na(x))})
```

```{r}
train_na_cols <- names(train_na_sums[train_na_sums > 0])
test_na_cols <- names(test_na_sums[test_na_sums > 0])

modelTrain <- modelTrain[, !names(modelTrain) %in% train_na_cols]
modelTest <- modelTest[, !names(modelTest) %in% test_na_cols]

str(modelTrain)
```

We can also eliminate the first seven rows since they do not deal with movement and would not help in predicting the classe variable.

```{r}
modelTrain <- modelTrain[, -c(1:7)]
modelTest <- modelTest[, -c(1:7)]
```

## Building the Model

Now we can train the model by using all remaining variables in the data set. I have chosen to use the random forest method since it is one of the most accurate methods. I am also going to save the model since it takes a while to build.

```{r}
# uncomment these when building the model
#model <- train(classe ~ ., method = "rf", data = modelTrain)
#saveRDS(model, "model.RDS")
model <- readRDS("model.RDS")
```

## Testing the Model

Now that we have built the model, we can test its accuracy on the testing data set.

```{r}
predictions <- predict(model, modelTest)
confusionMatrix(predictions, modelTest$classe)
```

As shown above, the model was **99.4%** accurate, so that means we estimate our out of sample error to be about **0.6%**. We can assume our model is quite accurate in predicting activity quality from activity monitors.








